---
title: "Lab2 Luciano Solution"
author: "LucianoRogerio"
date: "2022-01-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

<div id="menu" />


# Homework - Jan 31 2022

Illustrate regression to the mean between parents and offspring using AlphaSimR  
1. You know how to create a population -- create a population of parents. [Answer](#P1)  
2. You know how to get phenotypes from that population. [Answer](#P2)  
3. You want to randomly mate that population to get progeny that will be regressed to the mean. Create a `crossPlan` matrix like I did to generate the F1s, except that each row should have randomly-picked parents, rather than 1 and 2 like for the F1s. There is also a command `AlphaSimR::randCross`.  Check out its documentation. If you use that command, you will have to find the pedigree of the progeny using the `@mother` and `@father` ids of the progeny population and match those up to the parent population. If you make the crossPlan, then it gives you the seed and pollen parent ids.  [Answer](#P3)  
4. Having made the progeny population, phenotype it also.  [Answer](#P4)  
5. Use each row of the `crossPlan` to find the two parents and calculate their phenotypic mean.  [Answer](#P5)  
6. Make a scatterplot of the progeny phenotypes against the parent mean phenotypes.  [Answer](#P6)  

> What is the regression coefficient?

- Regression coefficient in this case is the broad sense heritability index of the simulated trait

> How much closer to the mean are offspring, on average, compared to their parents? 

- The offspring mean observed were 0.42 of the parent mean in one of my Simulations, but this value keeps changing if I run all again, even using the set.seed.
- I was expecting 0.5 that was the simulation parameter added to broad sense heritability, [results](#P7)

### Checking if the packages required are installed
```{r Checking installed packages, message = TRUE}
packages_used <- c("AlphaSimR", "tidyverse", "workflowr", "here")
ip <- installed.packages()
all_packages_installed <- TRUE
for (package in packages_used){
  if (!(package %in% ip[, "Package"])){
    print(paste("Please install package", package, sep = " "))
    all_packages_installed <- FALSE
  }
}#END packages_used
if (!all_packages_installed) stop("Need to install packages")

library(tidyverse); library(here)
```

loading the here package it already fix the Project directory as the working directory.

### Loading the packages and creating a README.Rmd to add information of software used in this analysis
```{r Load packages and README file, message = TRUE}
source(here::here("code/addToREADME.R"))
addToREADME(paste0("## ", rmarkdown::metadata$title), append = F)
addToREADME(c(date(), ""))

packages_info <- ip[packages_used, c("Package", "Version", "Built")]
addToREADME(c("The packages used in this script are:", "Package, Version, Built"))
apply(packages_info, 1, function(vec) addToREADME(paste(vec, collapse = " ")))
```

<div id="P1" />

### Seed set and Simulations parameters
```{r seed set and simulations parameters, message = TRUE}
set.seed(1312022)

addToREADME("")

addToREADME(paste("set.seed =", "1312022", sep = " "))
addToREADME("")

# Simulations parameters for the population
nFounders <- 100
inbred <- FALSE
ploidy <- 2L
nChr <- 18
effPopSize <- 200
segSites <- 1650
nQTL <- 1000
nSNP <- 500
nCores <- 4


addToREADME(c(paste("Number of Founders -->", nFounders, sep = " "),
              paste("Is the founders inbred?", inbred, sep = " "),
              paste("Which is the ploidy?", ploidy, sep = " "),
              paste("Number of Chromosomes -->", nChr, sep = " "),
              paste("Size of the effective population -->", effPopSize, sep = " "),
              paste("Number of quatitative loci traits per chromosome expected affecting the trait -->", nQTL, sep = " "),
              paste("Number of SNP markers per chromosome -->", nSNP, sep = " "),
              paste("Number of cores to apply parallel to the simulations of the chromosomes -->", nCores, sep = " ")))

addToREADME(" ")
```

<div id="P2" />


### Creating the Founder population and the parents population
```{r Founder and parent populations, message = TRUE}
founderPop <- AlphaSimR::runMacs(nInd = nFounders,
                                 inbred = inbred,
                                 ploidy = ploidy,
                                 nChr = nChr,
                                 segSites = segSites,
                                 nThreads = nCores)

addToREADME("Founder Pop Info")
addToREADME("----------------")
capture.output(founderPop, file = here::here("docs/README.md"), type = "output", append = TRUE)
addToREADME(" ")

SP <- AlphaSimR::SimParam$new(founderPop = founderPop)$addTraitA(nQtlPerChr = nQTL)$setVarE(H2 = 0.5)

# Simulating the Founder Pop with phenotype
foundersID <- AlphaSimR::newPop(rawPop = founderPop,
                                    simParam = SP)

addToREADME("Founders Ids Info")
addToREADME("----------------")
capture.output(foundersID, file = here::here("docs/README.md"), type = "output", append = TRUE)
addToREADME(" ")
```


<div id="P3" />

### Simulating the F1 and F2 populations
```{r 'Simulated F1 and F2 pop', message = TRUE}
F1 <- AlphaSimR::selectCross(pop = foundersID,
                             nInd = 35,
                             nCrosses = 200,
                             use = "pheno")
F1 <- AlphaSimR::setPheno(F1)
ParentsF1ID <- c(F1@mother,F1@father) %>% unique %>% .[order(as.numeric(.))]
ParentsF1 <- AlphaSimR::selectInd(foundersID, nInd = 35, candidates = ParentsF1ID)
ParentsF1 <- AlphaSimR::setPheno(ParentsF1)
F2 <- AlphaSimR::self(F1)
F2 <- AlphaSimR::setPheno(F2)

addToREADME("Populations Sim")
addToREADME("---------------")
addToREADME("Parents F1 Selected")
addToREADME("-------------------")
capture.output(ParentsF1, file = here::here("docs/README.md"), type = "output", append = TRUE)
addToREADME(" ")
addToREADME("F1 Simulated")
addToREADME("------------")
capture.output(F1, file = here::here("docs/README.md"), type = "output", append = TRUE)
addToREADME(" ")
addToREADME("F2 Simulated")
addToREADME("------------")
capture.output(F2, file = here::here("docs/README.md"), type = "output", append = TRUE)
addToREADME(" ")
```


<div id="P4" />

### Obtain the data from each Population
```{r Data collecting, message = TRUE}

ParentsF1Dt <- data.frame(ID = ParentsF1@id,
                          pheno = ParentsF1@pheno)

F1Dt <- data.frame(ID = F1@id,
                   mother = F1@mother,
                   father = F1@father,
                   pheno = F1@pheno)

F2Dt <- data.frame(ID = F2@id,
                   mother = F2@mother,
                   father = F2@father,
                   pheno = F2@pheno)
```

<div id="P5" />

### Prepare the Pop data with the Pheno of the parents in the same file
```{r Joint the Parents phenotypic data to the progeny dataset}
F1Dt2 <- left_join(F1Dt, ParentsF1Dt, by = c("mother" = "ID"), suffix = c("", ".mother")) %>%
  left_join(., ParentsF1Dt, by = c("father" = "ID"), suffix = c("", ".father")) %>%
  dplyr::mutate(pheno.ParMean = (pheno.mother + pheno.father)/2,
                Pop = "F1")

F2Dt2 <- left_join(F2Dt, ParentsF1Dt, by = c("mother" = "ID"), suffix = c("", ".mother")) %>%
  left_join(., ParentsF1Dt, by = c("father" = "ID"), suffix = c("", ".father")) %>%
  dplyr::mutate(pheno.ParMean = (pheno.mother + pheno.father)/2,
                Pop = "F2")

Data <- rbind(F1Dt2, F2Dt2)
```

<div id="P6" />

### Figure of the regression of the F1 and F2 progeny by it parent mean
```{r Figure, message = TRUE}
Data %>% 
  ggplot(aes(x = pheno.ParMean, y = pheno, color = Pop)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme_bw() + xlab("Parent Mean") + ylab("Progeny")
```

<div id="P7" />

### Estimation of the regression coefficient
```{r regression coefficient, message = TRUE}
Data %>% filter(Pop == "F1") %>%
  lm(formula = .$pheno ~ .$'pheno.ParMean') %>%
  coefficients %>% .[1] %>% round(., digits = 2) %>%
  paste("The observed broad sense heritability is", .,
        "while the expected was", SP$varG/(SP$varG + SP$varE), sep = " ")
```

[Menu](#menu)

[Home](LucSolutions.html)
